---
title: "Excess mortality during 2020 COVID-19 pandemic"
date: '`r format(Sys.Date(),"%B %e, %Y")`'
author: Aurora Hofman, Giovanni Rigolino, Paul Rognon
abstract: "During the COVID-19 pandemic obtaining the true number of the disease-related deaths is of critical value to evaluate the severity of the virus and public health responses. We investigate the relationship between the excess mortality observed in 2020 and reported COVID-19 deaths in four European countries. We conclude to the existence of excess deaths on top of COVID-19 reported deaths for several weeks and countries during the outbreak period."
output:
  pdf_document:
    number_sections: yes
    toc: yes
    includes:  
      in_header: my_header.tex
params:
  stanfile: "poison_time_dep.stan"
  iter: 5000
  adapt_delta: 0.99
  max_treedepth: 15
---
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, error = FALSE, message = FALSE,
  tidy.opts = list(width.cutoff = 55), eval = TRUE
)
```

```{r, echo=FALSE, eval = TRUE}
# libraries
library(rstan)
library(ggplot2)
library(tidyverse)
library(kableExtra)
library(tidyverse)
library(gridExtra)
library(bayesplot)
library(lubridate)
library(reshape)

# data
load("data_report.Rdata")
```

# Abstract

During the COVID-19 pandemic obtaining the true number of the disease-related deaths is of critical value to evaluate the severity of the virus and public health responses. We investigate the relationship between the excess mortality observed in 2020 and reported COVID-19 deaths in four European countries. We conclude to the existence of excess deaths on top of COVID-19 reported deaths for several weeks and countries during the outbreak period.

# Introduction

As of June 10th 2020, the global death toll of the unfolding COVID-19 outbreak stands at 413 126 [1]. The global and national official death toll figures have drawn much attention and sparked vivid debate because they are at the center of controversies of the severity of the virus and on the evaluation and comparison of the public health responses of governments. 

Among the controversies, there is first a debate on the severity of COVID-19 and many has compared it to the yearly flu outbreaks. Brazil's President Jair Bolsonaro referred the COVID-19 as a 'little flu' and refused to implement in his country the drastic lockdown measures that many other countries have enforced [2]. 

Second, the limits of testing in terms of tests availability and accuracy have led many observers to point out the likelihood of underreporting of deaths due to the novel virus. Moreover, deaths that might be indirectly due to the COVID-19 crisis because of, for example, a collapse of the health system, are not counted in official figures. On April 26th 2020, the Financial Times headlined that global coronavirus death toll could be 60% higher than reported [3]. 

Finally, differences in testing and reporting policies across countries but also within regions have casted more doubts on the veracity of the reported figures. Belgium has reported the highest number of deaths per 100 000 inhabitants [4] but Belgian officials also say they are counting in a way that no other country in the world is currently doing: counting deaths in hospitals and care homes, but also including deaths in care homes that are suspected, not confirmed, as COVID-19 cases [5].

There is a need of a rigorous estimation of the excess mortality in the weeks of the outbreak. A direct week by week comparison of the observed number of deaths to historical averages as done by the Financial Times analysts is a first approach but is limited as it fails to consider the variance of the number of deaths across years. 

In this report, we propose a Bayesian approach to estimate the excess mortality in the outbreak weeks through relative risk. Our model intends to provide parts of the answer to the following questions:

* Is mortality significantly higher than usual in the weeks of the outbreak?

* If confirmed, is there significant excess mortality on top of the reported COVID-19 deaths?

# Data

## Description

Every European state has an established monitoring system of death of nationals, often centralized by the local national statistics institute. Those platforms offer a good quality and reliable source to estimate excess mortality. We gathered weekly data of the total number of deaths for weeks 1 to 18  (end of April) in the following countries and following years:

* Norway: years 2014 to 2020 [6]

* Belgium: years 2009 to 2020 [7]

* France: years 2010 to 2020 [8]

* England and Wales: years 2010 to 2020 [9]

We picked those countries for different reasons: Belgium for its peculiar death counting methodology, France and England and Wales because they are among the most affected countries and Norway because it did not have a large outbreak and therefore offers a good benchmark. The data have been retrieved from the national statistic institutes of the respective countries. For each country the data were transformed to a similar structure as is visualized in Table \ref{tab:tab1}. 

The reported COVID-19 deaths for Norway, Belgium and France have been extracted from the Center for Systems Science and Engineering's repository at John Hopkins University's (JHU) Whiting School of Engineering [1]. The reported deaths for England and Wales comes from the Office of National Statistics (ONS) [9] as JHU did not have the count split by nations of the United Kingdom. However, we note that the JHU registered 28 446 COVID-19 deaths for the United Kingdom in week 18 [1], while ONS registered 36 595 [9]. The COVID-19 deaths figures for England and Wales are therefore more pessimistic than those of the other countries considered here.

```{r, echo=FALSE}
### Data

## loading weekly deaths data
data_england_wales <- read.csv("data/England_Wales_weekly.csv", sep = ",")
data_norway <- read.csv("data/Norway_total_death.csv", sep = ",")
data_france <- read.csv("data/weekly_deaths_france_week1_18_years2010_2020.csv", header = T, sep = ",")
data_belgium <- read.csv("data/weekly_deaths_belgium_week1_18_years2009_2020.csv", header = T, sep = ",")

## covid_data
data_covid <- read_csv("data/time_series_covid19_deaths_global.csv",
  col_types = cols(Lat = col_skip(), Long = col_skip())
)
data_covid <- data_covid %>%
  filter(`Country/Region` %in% c("Belgium", "France", "Norway"), is.na(`Province/State`)) %>%
  select(-`Province/State`)
colnames(data_covid)[1] <- "Country"
data_covid <- pivot_longer(data_covid, cols = contains("/"), names_to = "date", values_to = "total_deaths")
data_covid <- data_covid %>% mutate(date = parse_date_time(date, order = "mdy"), week = week(date))
data_covid <- data_covid %>%
  group_by(Country) %>%
  mutate(daily_death = c(0, diff(total_deaths)))
data_covid <- data_covid %>%
  group_by(Country, week) %>%
  mutate(weekly_death = sum(daily_death))
data_covid <- data_covid %>%
  select(Country, week, weekly_death) %>%
  distinct() %>%
  ungroup()
deaths_first_weeks <- data.frame(
  Country = c(rep("Belgium", 3), rep("Norway", 3), rep("France", 3)),
  week = rep(1:3, 3), weekly_death = rep(0, 9)
)
data_covid <- rbind(data_covid, deaths_first_weeks)
data_covid <- data_covid %>%
  filter(week < 19) %>%
  arrange(Country, week)

england_wales_covid <- read_csv("data/korona_england_wales.csv")
```

```{r, echo=FALSE}
## Extracting mean weekly deaths up to 2019, 2020 deaths and covid deaths by country

# England and Wales
mean_england <- rowMeans(data_england_wales[, paste("X", seq(2010, 2019), sep = "")])
england_2020 <- data_england_wales[, "X2020"]
england_covid <- england_wales_covid$england_wales

# Norway
mean_norway <- rowMeans(data_norway[, paste("X", seq(2014, 2019), sep = "")])
norway_2020 <- data_norway[, "X2020"]
norway_covid <- data_covid %>% filter(Country == "Norway")

# France
mean_france <- rowMeans(data_france[, paste("X", seq(2010, 2019), sep = "")])
france_2020 <- data_france[, "X2020"]
france_covid <- data_covid %>% filter(Country == "France")

# Belgium
mean_belgium <- rowMeans(data_belgium[, paste("X", seq(2009, 2019), sep = "")])
belgium_2020 <- data_belgium[, "X2020"]
belgium_covid <- data_covid %>% filter(Country == "Belgium")
```

```{r, echo = FALSE, eval = TRUE}
df <- data_england_wales
df <- df[, c("week", "X2017", "X2018", "X2019", "X2020")]
kable(df, "latex",
  booktabs = T,
  align = "c",
  col.names = c("Week", "2017", "2018", "2019", "2020"),
  caption = "\\label{tab:tab1} Number of deaths by week in England and Wales"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  )
```

## Visualization

### Weekly deaths

Fig.\ref{fig:fig1} shows the weekly deaths from weeks 1 to 18 for each of the countries and each of the years. We observe several common patterns across countries. Firstly, they all show a downward trend with larger numbers of deaths in the first weeks of the year than during spring. Secondly, there is noise and peaks around this trend with some years being more deadly than others. Earlier weeks are more subject to outbreaks of flu and other winter diseases than later weeks. We rarely observe dips in the series meaning the mortality has a higher tendency of increaced mortality than decreased mortality. This could be explained by a lower bound on mortality determined by the population age structure. Finally, most countries have a large peak between weeks 10 and 17 in 2020 which clearly stands out from the other years. On the contrary, Norway shows no peak.

```{r, echo = FALSE, eval = TRUE}
### Exploratory data analysis

# Weekly deaths across years
pivot_and_plot <- function(data) {
  data <- pivot_longer(data, cols = starts_with("X"), names_to = "year", values_to = "deaths")
  data$year <- gsub("X", "", data$year)
  data <- data %>% mutate(is2020=case_when(year < 2020 ~ "Prior to 2020",
                                           TRUE ~ "2020"))  
  p <- ggplot(data) +
    geom_line(aes(x = week, y = deaths, group = year, col = is2020)) +
    theme(legend.text = element_text(size = 4)) +
    guides(col = guide_legend(title = "Year")) +
    xlab("Week") +
    ylab("Deaths") +
    scale_color_manual(labels = c("2020","Prior to 2020"), values = c("red","grey"))
  return(p)
}

datasets <- list(
  data_england_wales,
  data_norway, data_france,
  data_belgium
)
countries <- c(
  "England and Wales", "Norway",
  "France", "Belgium"
)
plots <- list()

for (i in 1:length(datasets)) {
  plots[[i]] <- pivot_and_plot(datasets[[i]]) + ggtitle(countries[i])
}
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig1} Weekly deaths across years"}
grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]], ncol = 2)
```

### COVID-19 deaths and mean weekly deaths

In Fig.\ref{fig:fig2}, we compare the 2020 weekly deaths to the COVID-19 reported deaths summed to the mean weekly deaths. We observe for England and Wales a large difference between the two curves for weeks 13 to 18: there is excess mortality on top of expected deaths plus reported COVID-19 deaths. In Norway, 2020 deaths are below average up to week 13. After, the 2020 deaths are higher than mean deaths but still lower than COVID-19 and mean deaths together, suggesting a lower mortality in Norway in 2020 for non-COVID-19 related deaths which COVID-19 deaths do not compensate. Finally for Belgium and France, we observe larger 2020 deaths than mean deaths and COVID-19 together in weeks 11 to 15, suggesting again excess mortality on top of expected deaths plus reported COVID-19 deaths. However from week 16, the mean deaths plus COVID-19 curve is higher than 2020 deaths.

Overall in this quick exploration of our data, we find no issue or dubious data point.

```{r, echo = FALSE, eval = TRUE}
# 2020 weekly deaths, mean weekly deaths and COVID reported deaths
covid_mean_norway <- data.frame(week = seq(1:18), mean = mean_norway[1:18], `2020` = norway_2020[1:18], COVID = norway_covid$weekly_death)
covid_mean_france <- data.frame(week = seq(1:18), mean = mean_france[1:18], `2020` = france_2020[1:18], COVID = france_covid$weekly_death)
covid_mean_belgium <- data.frame(week = seq(1:18), mean = mean_belgium[1:18], `2020` = belgium_2020[1:18], COVID = belgium_covid$weekly_death)
covid_mean_england <- data.frame(week = seq(1:18), mean = mean_england[1:18], `2020` = england_2020[1:18], COVID = england_covid)

pivot_and_plot2 <- function(data) {
  data <- data %>% mutate(`mean+COVID` = mean + COVID)
  data <- pivot_longer(data, cols = c("mean", "mean+COVID", "X2020"), names_to = "type", values_to = "deaths")
  p <- ggplot(data) +
    geom_line(aes(x = week, y = deaths, col = type)) +
    scale_color_manual(labels = c("Mean deaths", "Mean deaths+\nCOVID", "2020 deaths"), values = c("green", "blue", "red")) +
    guides(col = guide_legend(title = "")) +
    xlab("Week") +
    ylab("Deaths")

  return(p)
}


datasets <- list(
  covid_mean_england,
  covid_mean_norway, covid_mean_france,
  covid_mean_belgium
)
countries <- c(
  "England and Wales", "Norway",
  "France", "Belgium"
)
plots2 <- list()

for (i in 1:length(datasets)) {
  plots2[[i]] <- pivot_and_plot2(datasets[[i]]) + ggtitle(countries[i])
}
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig2} 2020 weekly deaths, mean weekly deaths and COVID reported deaths"}
grid.arrange(plots2[[1]], plots2[[2]], plots2[[3]], plots2[[4]], ncol = 2)
```


# Model 

## Excess mortality

### Statistical model

For each country separately, we approach the modelling of excess mortality by estimating the relative risk $\gamma$ understood as the ratio of weekly mortality during the COVID-19 outbreak and the weekly mortality in non-outbreak times.

$$Relative Risk_i = \gamma_i = \frac{mortality_{COV,i}}{mortality_{noCOV,i}}=\frac{\frac{O_{COV,i}}{N}}{\frac{O_{noCOV,i}}{N}} = \frac{O_{COV,i}}{O_{noCOV,i}}$$

where $O_{COV,i}$ is the observed number of deaths in week $i$ of COVID-19 outbreak, $O_{noCOV,i}$ the number of deaths for the same week in non-outbreak times and $N$ is the population assumed stable across years.

The number of deaths $O$ is a count variable. For such variable, a commonly used statistical model is the Poisson model.

$$\forall i\in\{1,...,18\}, O_{COV,i} \sim \mathcal{PO}(\lambda), \hspace{0.1cm}E(O_{COV,i}) = Var(O_{COV,i}) = \lambda$$

We have:

$$\forall i\in\{1,...,18\},O_{COV,i}= O_{noCOV,i}\cdot\gamma_i \Rightarrow E(O_{COV,i})= E(O_{noCOV,i})\gamma_i = E_i \gamma_i$$

where $E_i$ is the expected number of deaths in week $i$ in non-outbreak times. For each week $i$, we estimate $E_i$  as the historical average of number of deaths in years prior to 2020.

We therefore define the following statistical model for each country separately:

$$\forall i\in\{1,...,18\},O_{COV,i} \sim \mathcal{PO}( E_i \gamma_i)$$
In these models, a value of $\gamma_i$ larger than 1 can be interpreted as excess mortality in week $i$ with respect to non-outbreak times and a value smaller than 1 as reduced mortality. This approach to modelling mortality is handy as the estimated value is standardized to the expected number of deaths and can therefore be compared between weeks.

From the exploratory data analysis, we see that there is a clear time dependency in the weekly mortality. We therefore decompose our relative risk $\gamma_i$ between a fixed component $\theta_i$ and a time structured component $\theta_{i,t}$: 
$$\gamma_i=\theta_{i}\theta_{t,i} \Rightarrow \log(\gamma_i)=\log(\theta_{i})+\log(\theta_{t,i})$$

We use an auto-regressive structure of order 1 as time structure. Our models then become:

$$\forall i\in\{1,...,18\},O_{COV,i} \sim \mathcal{PO}(E_i\gamma_i)= \mathcal{PO}(E_i\theta_i\theta_{t,i})$$

where:
$$\forall i\in\{2,...,18\},\log(\theta_{t,i}) = \alpha + \beta \log(\theta_{t,i-1}) $$

### Bayesian model

We are modelling the number of deaths $O_{COV,i}$ across weeks with a Poisson distribution that depends on the relative risk of each week $\gamma_i$. We are assuming different risks for each week but at the same time the risks share characteristics such as being expected at a value around 1 and having a larger probability towards higher values (peaks) than lower values (no trough). A convenient way to model this is using a common distribution for each of the components of $\gamma_i$, that is to use a Bayesian hierarchical model. We then propose a Bayesian hierarchical model based on the Poisson distribution for each country separately:

$$\forall i\in\{1,...,18\},(O_{COV,i}|\theta_i,\theta_{t,i}) \sim {\mathcal{PO}(E_i \theta_i\theta_{t,i})}$$

with parameters priors:

$$(\log(\theta_{1}|\sigma),...,\log(\theta_{18}|\sigma)) \sim \mathcal{N}(0,\sigma)$$

$$\forall i\in\{2,...,18\},\log(\theta_{t,i}|\sigma_t) = \alpha + \beta \log(\theta_{t,i-1}|\sigma_t) +\epsilon, \hspace{0.15cm} \epsilon \sim \mathcal{N}(0,\sigma_t)$$


where $\sigma$, $\alpha$, $\beta$ and $\sigma_t$ are hyperparameters with hyperpriors:

$$\sigma \sim \mathcal{U}(a_{\sigma},b_{\sigma})$$
$$\alpha \sim \mathcal{N}(\mu_{\alpha},\sigma_{\alpha})$$
$$\beta \sim \mathcal{U}(a_{\beta},b_{\beta})$$
$$\sigma_t \sim \mathcal{U}(a_{\sigma_t},b_{\sigma_t})$$

## Excess mortality on top of the reported COVID-19 deaths

To model excess mortality on top of reported COVID-19 deaths, we use the same approach as for the modelling of excess mortality during the COVID-19 outbreak with respect to non-outbreak times. We previously estimated the relative risk $\gamma$ understood as the ratio of weekly mortality during the COVID-19 outbreak and the weekly expected mortality in non-outbreak times. We now adjust, for each country separately, the expected mortality $\tilde{E_i}$ to account for COVID-19 reported deaths.

$$Relative Risk_i = \gamma_i = \frac{O_{COV,i}}{O_{noCOV,i}+D_i}$$

where $O_{COV,i}$ is the observed number of deaths in week $i$ of COVID-19 outbreak, $O_{noCOV,i}$ the number of deaths for the same week in non-outbreak times and $D_i$ is the reported number of deaths due to COVID-19 in week $i$.

We have:

$$\tilde{E_i}=E_i+D_i$$

The updated statistical model is:

$$\forall i\in\{1,...,18\},O_{COV,i} \sim \mathcal{PO}(\tilde{E}_i\gamma_i)= \mathcal{PO}(\tilde{E}\theta_i\theta_{t,i})$$

where:
$$\forall i\in\{2,...,18\},\log(\theta_{t,i}) = \alpha + \beta \log(\theta_{t,i-1}) $$
The priors and hyperpriors for the Bayesian model do not change.

### Choice of priors

Our choice of priors and hyperpriors has been driven by prior knowledge, or lack of it. First of all, we choose a normal distribution for $\log(\theta_i)$ with mean 0 and standard deviation the hyperparameter $\sigma$. As a result $\theta_i$ is lognormally distributed and centered at 1, the expected value for the relative risk. The lognormal distribution is particularly suited for several reasons. It has a positive support and we know $\theta_i$ must be positive. The lognormal distribution has also a positive skew, so the median of $\theta_i$ is 1 but the mean is larger and very large values are possible. It is important to have such heavy right tail because we know the risk can increase largely when large scale deadly events occur. On the contrary, we do not need a heavy left tail because the mortality is floored by the natural mortality determined by the population demographics. For the time dependent component of relative risk, we make the same distributional assumption where the center of the distribution is not 1 anymore but the autoregression function estimate $\alpha+\beta\theta_{t,i-1}$, that represents the dependency on previous value.

#### Choice of hyperpriors

The choice of a uniform distribution with support $(-1,1)$ for $\beta$ is motivated by our assumption of autoregressive model of order 1 for $\log(\theta_{t,i})$. When the autoregresive process is stationary, $\beta$ is the autocorrelation between two consecutive values and must be between -1 and 1. The choice of the non-informative uniform distribution reflects our lack of further knowledge on the distribution of $\beta$. We choose for $\alpha$ the normal distribution, commonly used in linear regression settings. We assume the distribution of $\alpha$ to be centered at 0 because in a stationary autoregressive model of order 1, a null constant term $\alpha$ makes a null expectation for the modelled variable:

$$\log(\theta_{t,i}) = \alpha + \beta \log(\theta_{t,i-1}) \Rightarrow E(\log(\theta_{t,i})) = \alpha + \beta E(\log(\theta_{t,i-1})) \Rightarrow \mu = \alpha + \beta \mu \Rightarrow \mu=\frac{\alpha}{1-\beta}$$

As our modelled variable $\theta_{t,_i}$ is log transformed, it converts to an expectation of 1 when untransformed. We choose a small value for $\sigma_{\alpha}$, 0.2, to prevent large deviations towards extremely high risk.

Finally, we choose non-informative uniform distributions for $\sigma$ and $\sigma_t$ because we do not hold prior information on them. We define small supports, $(0,0.2)$, to again prevent large deviations towards extremely high risks, as exemplified below.

```{r}
set.seed(42)
#### Prior predictive for gamma = theta * theta_t

# hyperparameters
sigma_a <- 0
sigma_b <- 0.2
alpha_mu <- 0
alpha_sigma <- 0.2
beta_a <- -1
beta_b <- 1
sigma_time_a <- 0
sigma_time_b <- 0.2

# simulation function
prior_pred_theta <- function() {
  sigma <- runif(1, sigma_a, sigma_b)
  while (sigma <= 0.01) {
    sigma <- runif(1, sigma_a, sigma_b)
  }
  # sigma_time <- runif(1, 0.01, 25)

  sigma_time <- runif(1, sigma_time_a, sigma_time_b)
  while (sigma_time <= 0.01) {
    sigma_time <- runif(1, sigma_time_a, sigma_time_b)
  }

  alpha <- rnorm(1, mean = alpha_mu, sd = alpha_sigma)
  beta <- runif(1, beta_a, beta_b)
  log_theta <- rnorm(18, mean = 0, sigma)
  log_theta_time <- rep(0, 18)
  log_theta_time[1] <- rnorm(1, mean = 0, sigma_time)
  for (i in 2:18) {
    log_theta_time[i] <- rnorm(1, mean = alpha + beta * log_theta_time[i - 1], sigma_time)
  }
  theta_return <- exp(log_theta + log_theta_time)

  return(theta_return)
}

# simulation resuls
matrix_theta <- c()
for (i in 1:30000) {
  matrix_theta <- c(matrix_theta, prior_pred_theta())
}
mean_priorpred_risk <- mean(matrix_theta)
median_priorpred_risk <- median(matrix_theta)
q095_priorpred_risk <- quantile(matrix_theta, 0.95)
q005_priorpred_risk <- quantile(matrix_theta, 0.05)
```

We check our prior by visualizing the prior predictive distribution of the relative risk and the number of deaths. We first simulated the prior predictive distribution of the total relative risk $\gamma$ across all weeks and present the results in Fig.\ref{fig:fig3}. We see the obtained distribution is concentrated around 1, the expected value for the risk, with the majority of the density between 0.5 and 1.5, a reasonable range of values for the risk. The quantile 5% stands at `r round(q005_priorpred_risk,3)` and the quantile 95% at `r round(q095_priorpred_risk,3)`. The right tail is heavy which is appropriate to capture peaks. Wider supports for the hyperpriors of $\sigma$, $\sigma_{\alpha}$ and $\sigma_t$ yielded to too heavy right tails.  

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig3} Density of prior predictive distribution of risk "}
# plot
matrix_theta_filt <- data.frame(risk = matrix_theta) %>% filter(risk <= quantile(matrix_theta, 0.99))
ggplot(matrix_theta_filt, aes(risk)) +
  geom_density() +
  xlab("Total relative risk (truncated at q99%)") +
  theme_light() +
  ylab("Density")
```

We simulated the prior predictive distribution of the number of deaths for weeks 1 to 18 and plot the results in Fig.\ref{fig:fig4} for each country. We see the obtained distributions are roughly centered around the historical mean up to 2019, the observed mean is very close to the distributions medians. For every week, the range of values is wide, allowing for large but reasonable variations of the number of deaths. In accordance with the objective of accommodating peaks, there is more room for variation away from the center towards higher values.

```{r}
#### Prior predictive for number of deaths

# Simulation function
prior_pred_o <- function(mean_data) {
  sigma <- runif(1, sigma_a, sigma_b)
  while (sigma <= 0.01) {
    sigma <- runif(1, sigma_a, sigma_b)
  }

  sigma_time <- runif(1, sigma_a, sigma_b)
  while (sigma_time <= 0.01) {
    sigma_time <- runif(1, sigma_a, sigma_b)
  }

  alpha <- rnorm(1, mean = alpha_mu, sd = alpha_sigma)
  beta <- runif(1, beta_a, beta_b)
  log_theta <- rnorm(length(mean_data), mean = 0, sigma)
  log_theta_time <- rep(0, length(mean_data))
  log_theta_time[1] <- rnorm(1, mean = 0, sigma_time)
  for (i in 2:length(mean_data)) {
    log_theta_time[i] <- rnorm(1, mean = alpha + beta * log_theta_time[i - 1], sigma_time)
  }
  theta_return <- exp(log_theta + log_theta_time)
  O <- rep(0, length(mean_data))
  for (i in 1:length(mean_data)) {
    O[i] <- rpois(1, mean_data[i] * theta_return[i])
  }
  return(O)
}
```

```{r}
# Plot function
plot_priorpred_o <- function(mean) {
  matrix_O <- c()

  for (i in 1:10000) {
    matrix_O <- rbind(matrix_O, prior_pred_o(mean))
  }
  matrix_O <- data.frame(matrix_O)

  df <- matrix_O %>% pivot_longer(cols = starts_with("X"), names_to = "week")

  df <- df %>%
    group_by(week) %>%
    summarise(
      q_0.05 = quantile(value, 0.05),
      median_o = median(value),
      q_0.95 = quantile(value, 0.95)
    ) %>%
    ungroup()
  df <- df %>%
    mutate(week = as.numeric(gsub("X", "", df$week))) %>%
    arrange(week) %>%
    mutate(obs_mean = mean)

  df <- df %>%
    pivot_longer(cols = c(obs_mean, q_0.05, q_0.95, median_o), names_to = "metric")

  p <- ggplot(data = df) +
    geom_line(aes(x = week, y = value, linetype = metric, col = metric)) +
    scale_linetype_manual(name = "", values = c(1, 1, 2, 2), labels = c("Obs. mean", "Sim. median", "Sim. q5%", "Sim. q95%")) +
    scale_color_manual(name = "", values = c("blue", "black", "red", "red"), labels = c("Obs. mean", "Sim. median", "Sim. q5%", "Sim. q95%")) +
    theme_light() +
    theme(legend.position = "bottom") +
    xlab("Week") +
    ylab("Deaths")

  return(p)
}

datasets_mean <- list(
  mean_england,
  mean_norway, mean_france,
  mean_belgium
)
countries <- c(
  "England and Wales", "Norway",
  "France", "Belgium"
)
plots3 <- list()

for (i in 1:length(datasets_mean)) {
  plots3[[i]] <- plot_priorpred_o(datasets_mean[[i]]) + ggtitle(countries[i])
}
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig4} Prior predictive distribution of number of deaths"}
grid.arrange(plots3[[1]], plots3[[2]], plots3[[3]], plots3[[4]], ncol = 2)
```

# Results

```{r}
### STAN runs

# run in Google Colab with attached notebooks

## Excess mortality stan

# params lists
create_list <- function(N, E, O) {
  data_list <- list(
    N = N,
    E = E,
    O = O,
    sigma_a = 0,
    sigma_b = 0.2,
    alpha_mu = 0,
    alpha_sigma = 0.2,
    beta_a = -1,
    beta_b = 1,
    sigma_time_a = 0,
    sigma_time_b = 0.2
  )
  return(data_list)
}

data_list_england_wales <- create_list(length(mean_england), mean_england, england_2020)
data_list_norway <- create_list(length(mean_norway), mean_norway, norway_2020)
data_list_france <- create_list(length(mean_france), mean_france, france_2020)
data_list_belgium <- create_list(length(mean_belgium), mean_belgium, belgium_2020)
```

```{r, eval = FALSE}
load("data_for_stan.RData")

# stan run. run in Google Colab
risk_england_wales <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_england_wales, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)

risk_norway <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_norway, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)

risk_france <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_france, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)

risk_belgium <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_belgium, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)
```


```{r}
## Excess mortality on top of the reported COVID-19 deaths stan
# params lists

data_list_england_wales_covid <- create_list(
  length(mean_england),
  mean_england + england_covid,
  england_2020
)
data_list_norway_covid <- create_list(
  length(mean_norway),
  mean_norway + norway_covid$weekly_death,
  norway_2020
)
data_list_france_covid <- create_list(
  length(mean_france),
  mean_france + france_covid$weekly_death,
  france_2020
)
data_list_belgium_covid <- create_list(
  length(mean_belgium),
  mean_belgium + belgium_covid$weekly_death,
  belgium_2020
)
save.image("data_for_stan.RData")
```

```{r, eval = FALSE}
load("data_for_stan.RData")

# stan run. run in Google Colab
risk_england_wales_covid <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_england_wales_covid, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)

risk_norway_covid <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_norway_covid, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)


risk_france_covid <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_france_covid, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)

risk_belgium_covid <- stan(params$stanfile,
  iter = params$iter, chains = 3,
  data = data_list_belgium_covid, seed = 1, control = list(adapt_delta = params$adapt_delta, max_treedepth = params$max_treedepth)
)
```


## Convergence

### Excess deaths
    
We check the convergence of each of our four models (one per country) by checking the Rhat values for all parameters, plotting the chains traces and autocorrelation in chains. Table \ref{tab:tab3} gathers the maximum Rhat value in each model. All the values are below 1.01 indicating satisfactory convergence. In Fig.\ref{fig:fig5} and Fig. \ref{fig:fig6} we plot the chain traces and autocorrelation plots for four parameters in the England and Wales model as examples. We can see the chains mix well. There is a bit of autocorrelation in the plots of log_theta[16] but it is small and there is no autocorrelation for the other parameters.

```{r,eval=FALSE}
### Convergence check 

## Excess deaths model

print(risk_england_wales)
print(risk_belgium)
print(risk_france)
print(risk_norway)
```

```{r}
df <- data.frame(
  country = c("England and Wales", "Norway", "Belgium", "France"),
  Max_rhat = c(
    max(rhat(risk_england_wales)),
    max(rhat(risk_norway)),
    max(rhat(risk_belgium)),
    max(rhat(risk_france))
  )
)
kable(df, "latex",
  booktabs = T,
  align = "c",
  digit = 3,
  col.names = c("Country", "Max rhat"),
  caption = "\\label{tab:tab3} Max Rhat in excess deaths models"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  )
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig5} Chains traces for a sample of parameters from England and Wales excess deaths model"}
# England and Wales trace
mcmc_trace(risk_england_wales,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig6} Autocorrelation plots for a sample of parameters from England and Wales excess deaths model"}
# England and Wales ACF
mcmc_acf_bar(risk_england_wales,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)
```

```{r, eval=FALSE}
# Norway
mcmc_acf_bar(risk_norway,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)

mcmc_trace(risk_norway,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)

# Belgium
mcmc_acf_bar(risk_belgium,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)

mcmc_trace(risk_belgium,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)

# France
mcmc_acf_bar(risk_france,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)

mcmc_trace(risk_france,
  pars = c("log_theta[16]", "log_theta_time[7]", "alpha", "sigma_time")
)
```


### Excess deaths on top of the reported COVID-19 deaths
    
We repeat the procedure for the model on excess deaths on top of the reported COVID-19 deaths. We check again the convergence of each of our four models (one per country) by checking the Rhat values for all parameters, plotting the chains traces and autocorrelations in chains. Table \ref{tab:tab4} gathers the maximum Rhat value in each model. All the values are below 1.02 indicating satisfactory convergence. In Fig.\ref{fig:fig7} and \ref{fig:fig8} we plot the chain traces and autocorrelation plots for four parameters in the Belgium model as examples. We can see the chains mix well. There is a bit of autocorrelation in the plots of log_theta[17] and beta but only in one chain and there is no autocorrelation for the other parameters.

```{r,eval=FALSE}
## Excess deaths on top of COVID reported deaths model

print(risk_england_wales_covid)
print(risk_belgium_covid)
print(risk_france_covid)
print(risk_norway_covid)
```

```{r}
df <- data.frame(
  country = c("England and Wales", "Norway", "Belgium", "France"),
  Max_rhat = c(
    max(rhat(risk_england_wales_covid)),
    max(rhat(risk_norway_covid)),
    max(rhat(risk_belgium_covid)),
    max(rhat(risk_france_covid))
  )
)
kable(df, "latex",
  booktabs = T,
  align = "c",
  digit = 3,
  col.names = c("Country", "Max rhat"),
  caption = "\\label{tab:tab4} Max Rhat in models for excess deaths on top of the reported COVID-19 deaths"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  )
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig7} Chains traces for a sample of parameters from the Belgium model on excess deaths on top of the reported COVID-19 deaths"}
# England and Wales trace
mcmc_trace(risk_belgium_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig8} Autocorrelation plots for a sample of parameters from the Belgium model on excess deaths on top of the reported COVID-19 deaths"}
# England and Wales ACF
mcmc_acf_bar(risk_belgium_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)
```

```{r, eval=FALSE}
# Norway
mcmc_acf_bar(risk_norway_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)

mcmc_trace(risk_norway_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)

# England and Wales
mcmc_acf_bar(risk_england_wales_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)

mcmc_trace(risk_england_wales_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)

# France
mcmc_acf_bar(risk_france_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)

mcmc_trace(risk_france_covid,
  pars = c("log_theta[1]", "log_theta_time[17]", "beta", "sigma_time")
)
```

## Validation

To validate our model we need to compare statistics from the posterior distribution to actual values. The only actual data we have is the number of deaths per week in 2020. We then validate our model working on the posterior predictive distribution of the number of deaths. For each country and for each week, we compare the 2020 value to the median and quantiles 5% and 95% of our posterior predictive distribution. In this validation procedure, a good model would be a model for which the 2020 data falls between the two quantiles for all weeks and close to the median. The 5% and 95% quantiles should also be reasonable values, less than 3 times the number of observed deaths and more than a third of it.

```{r}
### Model validation

# function for extracting simulated data
extract_simulations <- function(risk_data) {
  sim_1 <- rstan::extract(risk_data, "sim[1]")[[1]]
  sim_2 <- rstan::extract(risk_data, "sim[2]")[[1]]
  sim_3 <- rstan::extract(risk_data, "sim[3]")[[1]]
  sim_4 <- rstan::extract(risk_data, "sim[4]")[[1]]
  sim_5 <- rstan::extract(risk_data, "sim[5]")[[1]]
  sim_6 <- rstan::extract(risk_data, "sim[6]")[[1]]
  sim_7 <- rstan::extract(risk_data, "sim[7]")[[1]]
  sim_8 <- rstan::extract(risk_data, "sim[8]")[[1]]
  sim_9 <- rstan::extract(risk_data, "sim[9]")[[1]]
  sim_10 <- rstan::extract(risk_data, "sim[10]")[[1]]
  sim_11 <- rstan::extract(risk_data, "sim[11]")[[1]]
  sim_12 <- rstan::extract(risk_data, "sim[12]")[[1]]
  sim_13 <- rstan::extract(risk_data, "sim[13]")[[1]]
  sim_14 <- rstan::extract(risk_data, "sim[14]")[[1]]
  sim_15 <- rstan::extract(risk_data, "sim[15]")[[1]]
  sim_16 <- rstan::extract(risk_data, "sim[16]")[[1]]
  sim_17 <- rstan::extract(risk_data, "sim[17]")[[1]]
  sim_18 <- rstan::extract(risk_data, "sim[18]")[[1]]
  posterior_chains <- tibble(sim_1, sim_2, sim_3, sim_4, sim_5, sim_6, sim_7, sim_8, sim_9, sim_10, sim_11, sim_12, sim_13, sim_14, sim_15, sim_16, sim_17, sim_18)
  return(posterior_chains)
}

# function for validation plot
plot_validation <- function(posterior_chain, deaths_2020, lim = 14) {
  set.seed(43)
  df0 <- posterior_chain %>% pivot_longer(cols = starts_with("sim"), names_to = "week")

  df0 <- df0 %>%
    group_by(week) %>%
    summarise(
      Simq0.05 = quantile(value, 0.05),
      median_o = median(value),
      Simq0.95 = quantile(value, 0.95)
    ) %>%
    ungroup()

  df0 <- df0 %>%
    mutate(week = as.numeric(gsub("sim_", "", df0$week))) %>%
    arrange(week) %>%
    mutate(obs2020 = deaths_2020, bef_af = as.factor(week > lim))

  df <- df0 %>% pivot_longer(cols = c(obs2020, Simq0.05, Simq0.95), names_to = "metric")

  levels(df$bef_af) <- c(paste0("Weeks 1-", lim), paste0("Weeks ", lim + 1, "-18"))

  posterior_chains_mini <- data.frame(t(posterior_chain[sample(1:nrow(posterior_chain), 10), ]),
    week = seq(1:18), bef_af = as.factor(seq(1:18) > lim)
  )
  Molten <- melt(posterior_chains_mini, id.vars = c("week", "bef_af"))
  Molten$Simulation <- gsub("X", "", Molten$variable)

  levels(Molten$bef_af) <- c(paste0("Weeks 1-", lim), paste0("Weeks ", lim + 1, "-18"))

  p <- ggplot() +
    geom_line(data = Molten, aes(x = week, y = value, colour = Simulation)) +
    geom_line(data = df, aes(x = week, y = value, linetype = metric), col = "black", size = 0.7) +
    scale_linetype_manual(name = "", values = c(1, 2, 2), labels = c("Obs. 2020", "Sim. q5%", "Sim. q95%")) +
    facet_wrap(~bef_af, scales = "free") +
    theme_light() +
    theme(legend.position = "bottom") +
    xlab("Week") +
    ylab("Deaths")
  return(list(plot = p, comp_med = df0 %>% select(c(obs2020, median_o))))
}
```


### Excess deaths

In Fig.\ref{fig:fig9} and Fig.\ref{fig:fig9b} we plot, for England and Wales and Norway, the 2020 deaths along with 10 of the 6500 simulations of the posterior predictive distribution and the 5% and 95% quantiles computed on the simulated posterior predictive distribution for each week.The same plots are available in the Annex for Belgium and France. We find that, for all the countries, the 2020 deaths data fall well between the quantiles and in the mid of the simulated trajectories. The quantile values are also reasonable related to the criteria described before, maybe even too close to the actual values. In the Annex Tables \ref{tab:tab5} and \ref{tab:tab6} the 2020 deaths are compared to the median of the simulations of the posterior predictive for each week. We find that for all the countries, the two values are very close.

```{r}
## Excess deaths model

posterior_chains_england <- extract_simulations(risk_england_wales)
posterior_chains_norway <- extract_simulations(risk_norway)
posterior_chains_france <- extract_simulations(risk_france)
posterior_chains_belgium <- extract_simulations(risk_belgium)

val_plot_england <- plot_validation(posterior_chains_england, england_2020, lim = 14)$plot +
  ggtitle("England and Wales")

val_plot_belgium <- plot_validation(posterior_chains_belgium, belgium_2020, lim = 13)$plot +
  ggtitle("Belgium")

val_plot_france <- plot_validation(posterior_chains_france, france_2020, lim = 13)$plot + ggtitle("France")


val_plot_norway <- plot_validation(posterior_chains_norway, norway_2020, lim = 13)$plot + ggtitle("Norway")
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig9} Validation of excess deaths model"}
val_plot_england
```


```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig9b} Validation of excess deaths model continued"}
val_plot_norway
```

### Excess deaths on top of the reported COVID-19 deaths

We apply the same procedure to the model for excess deaths on top of the reported COVID-19 deaths. In Fig.\ref{fig:fig10c} and Fig. \ref{fig:fig10d} we plot, for Belgium and France, the 2020 deaths along with 10 simulations of the 6500 simulations of the posterior predictive distribution and the 5% and 95% quantiles computed on the posterior predictive for each week. The same plots are available for England and Wales and Norway in the Annex. We reach the same conclusions as for the excess deaths model above: 2020 deaths data fall well between the quantiles, quantile values are reasonable given the criteria described before. In the Annex, in Tables \ref{tab:tab7} and \ref{tab:tab8}, we compare the 2020 deaths to the median of the simulations of the posterior predictive for each week and again we observe that for all the countries the two values are very close.

```{r}
## Excess deaths model on top of the reported COVID-19 deaths
posterior_chains_england_covid <- extract_simulations(risk_england_wales_covid)
posterior_chains_norway_covid <- extract_simulations(risk_norway_covid)
posterior_chains_france_covid <- extract_simulations(risk_france_covid)
posterior_chains_belgium_covid <- extract_simulations(risk_belgium_covid)

val_plot_england_covid <- plot_validation(posterior_chains_england_covid, england_2020, lim = 14)$plot +
  ggtitle("England and Wales")

val_plot_belgium_covid <- plot_validation(posterior_chains_belgium_covid, belgium_2020, lim = 13)$plot +
  ggtitle("Belgium")

val_plot_france_covid <- plot_validation(posterior_chains_france_covid, france_2020, lim = 13)$plot + ggtitle("France")


val_plot_norway_covid <- plot_validation(posterior_chains_norway_covid, norway_2020, lim = 13)$plot + ggtitle("Norway")
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig10c} Validation of excess deaths on top of the reported COVID-19 deaths model continued"}
val_plot_belgium_covid
```


```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig10d} Validation of excess deaths on top of the reported COVID-19 deaths modelcontinued"}
val_plot_france_covid
```

## Inference

Now that we have confirmed the convergence and validity of our model, we are ready to provide estimates of the total relative risk $\gamma$ for every week in every country.

```{r,eval=FALSE}
### Inference

# Posterior distribution of selected parameters
p1 <- mcmc_areas(risk_england_wales,
  prob = 0.90,
  pars = c("log_theta[1]", "log_theta_time[13]", "log_theta_time[14]", "log_theta_time[18]")
) + ggtitle("England and Wales")
p2 <- mcmc_areas(risk_belgium,
  prob = 0.90,
  pars = c("log_theta[1]", "log_theta_time[12]", "log_theta_time[14]", "log_theta_time[18]")
) + ggtitle("Belgium")
p3 <- mcmc_areas(risk_france,
  prob = 0.90,
  pars = c("log_theta[1]", "log_theta_time[12]", "log_theta_time[14]", "log_theta_time[18]")
) + ggtitle("France")
p4 <- mcmc_areas(risk_norway,
  prob = 0.90,
  pars = c("log_theta[1]", "log_theta_time[12]", "log_theta_time[14]", "log_theta_time[18]")
) + ggtitle("Norway")
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

```{r}
# helper functions for extracting relative risk results
create_CI_theta_vec <- function(risk_data, N) {
  fit <- summary(risk_data)
  results <- as.data.frame(fit$summary)

  CI_upper <- exp(results$`97.5%`[1:N] + results$`97.5%`[(N + 1):(2 * N)])
  CI_lower <- exp(results$`2.5%`[1:N] + results$`2.5%`[(N + 1):(2 * N)])

  theta <- exp(results$mean[1:N] + results$mean[(N + 1):(2 * N)])
  week <- seq(1:N)
  data <- data.frame(week, CI_upper, CI_lower, theta)
  return(data)
}
plot_risk <- function(data, country_name) {
  g <- ggplot(data = data, aes(x = week, y = theta)) +
    geom_point() +
    geom_line(lty = 2) +
    geom_line(aes(week, CI_lower), col = "blue", lty = 2) +
    geom_line(aes(week, CI_upper), col = "blue", lty = 2) +
    geom_hline(yintercept = 1, col = "red") +
    xlab("Week") +
    ylab("Total relative risk") +
    ggtitle(country_name)
  #   ggsave(paste0("output_",country_name,".png"))
  return(g)
}
```


### Excess number of deaths

In Fig.\ref{fig:fig11} we plot the relative risk point estimates (mean) as well as their 95% credible intervals for the excess deaths models. We find that for England and Wales, France and Belgium we have several weeks for which the relative risk is estimated well above 1 and with a probability of being above 1 superior to 95%. Those weeks are weeks 12 to 17 for France and Belgium and 14 to 18 for England & Wales. Before those periods, the relative risk is very close to one and the 95% credible intervals do not exclude a neutral risk of 1. For Norway, all the relative risks are estimated close to 1 and 95% credible intervals exclude a neutral risk of 1 in none of the weeks considered.

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig11} Total relative risk estimates in excess deaths models"}
## Relative risk in excess deaths model
data_england_wales <- create_CI_theta_vec(risk_england_wales, length(mean_england))
risk1 <- plot_risk(data_england_wales, "England and Wales")

data_norway <- create_CI_theta_vec(risk_norway, length(mean_norway))
risk2 <- plot_risk(data_norway, "Norway")


data_france <- create_CI_theta_vec(risk_france, length(mean_france))
risk3 <- plot_risk(data_france, "France")

data_belgium <- create_CI_theta_vec(risk_belgium, length(mean_belgium))
risk4 <- plot_risk(data_belgium, "Belgium")
grid.arrange(risk1, risk2, risk3, risk4, ncol = 2)
```

### Excess deaths on top of the reported COVID-19 deaths

In Fig.\ref{fig:fig12} we plot the point estimates  (mean) of the relative risk for excess deaths on top of the reported COVID-19 deaths and their 95% credible intervals. We find that in all the countries but Norway, there are weeks where the relative risk is estimated well above 1 and with a probability of being above 1 superior to 95%. Those weeks are weeks 12 and 13 in France, 12 to 14 in Belgium and 13 to 18 in England & Wales. For Norway, all the relative risks are estimated close to 1 and 95% credible intervals exclude a neutral risk of 1 in none of the weeks considered. In Belgium and France the relative risk is estimated below 1 for weeks 15 to 18.

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig12} Total relative risk estimates in models for excess deaths on top of the reported COVID-19 deaths"}
## Relative risk in excess deaths on top of COVID deaths model

data_england_wales <- create_CI_theta_vec(risk_england_wales_covid, length(mean_england))
risk1_covid <- plot_risk(data_england_wales, "England and Wales")

data_norway <- create_CI_theta_vec(risk_norway_covid, length(mean_norway))
risk2_covid <- plot_risk(data_norway, "Norway")


data_france <- create_CI_theta_vec(risk_france_covid, length(mean_france))
risk3_covid <- plot_risk(data_france, "France")

data_belgium <- create_CI_theta_vec(risk_belgium_covid, length(mean_belgium))
risk4_covid <- plot_risk(data_belgium, "Belgium")
grid.arrange(risk1_covid, risk2_covid, risk3_covid, risk4_covid, ncol = 2)
```

With our model we can use the posterior predictive distribution of the number of deaths to provide an estimate of the excess deaths on top of the expected number of deaths plus the reported COVID-19 deaths. In In Fig.\ref{fig:fig13a} to \ref{fig:fig13d} we plot a point estimate (median) and 95% credible interval for the number of deaths and compare it to the expected number of deaths plus the reported COVID-19 death. In England and Wales, from week 11 to 18, the number of deaths is estimated well above the expected number of deaths plus the reported COVID-19 death, with a probability of being superior larger than 95%. The same occurs for weeks 12 to 14 in Belgium and 11 to 13 in France. In both countries, for the weeks after, the estimate is lower than the expected number of deaths plus the reported COVID-19 death. In Norway, all along the time span considered, the expected number of deaths plus the reported COVID-19 death falls within the 95% credible interval for the number of deaths.

```{r}
# function to plot posterior pred outputs for number of deaths
plot_excessdeaths <- function(posterior_chain, mean_plus_covid, lim = 14) {
  set.seed(43)
  df0 <- posterior_chain %>% pivot_longer(cols = starts_with("sim"), names_to = "week")

  df0 <- df0 %>%
    group_by(week) %>%
    summarise(
      Simq0.05 = quantile(value, 0.05),
      Simmedian = median(value),
      Simq0.95 = quantile(value, 0.95)
    ) %>%
    ungroup()

  df0 <- df0 %>%
    mutate(week = as.numeric(gsub("sim_", "", df0$week))) %>%
    arrange(week) %>%
    mutate(meanpluscovid = mean_plus_covid, bef_af = as.factor(week > lim))

  df <- df0 %>% pivot_longer(cols = c(meanpluscovid, Simmedian, Simq0.05, Simq0.95), names_to = "metric")

  levels(df$bef_af) <- c(paste0("Weeks 1-", lim), paste0("Weeks ", lim + 1, "-18"))

  p <- ggplot() +
    geom_line(data = df, aes(x = week, y = value, linetype = metric, col = metric), size = 0.7) +
    scale_linetype_manual(name = "", values = c(1, 1, 2, 2), labels = c("Mean + COVID", "Sim. median", "Sim. q5%", "Sim. q95%")) +
    scale_color_manual(name = "", values = c("blue", "green", "red", "red"), labels = c("Mean + COVID", "Sim. median", "Sim. q5%", "Sim. q95%")) +
    facet_wrap(~bef_af, scales = "free") +
    theme_light() +
    theme(legend.position = "bottom") +
    xlab("Week") +
    ylab("Deaths")
  return(list(plot = p, comp = df0 %>% select(-bef_af)))
}

excess_plot_england_covid <- plot_excessdeaths(posterior_chains_england_covid, mean_england + england_covid, lim = 14)$plot +
  ggtitle("England and Wales")

excess_plot_belgium_covid <- plot_excessdeaths(posterior_chains_belgium_covid, mean_belgium + belgium_covid$weekly_death, lim = 13)$plot + ggtitle("Belgium")

excess_plot_france_covid <- plot_excessdeaths(posterior_chains_france_covid, mean_france + france_covid$weekly_death, lim = 12)$plot + ggtitle("France")

excess_plot_norway_covid <- plot_excessdeaths(posterior_chains_norway_covid, mean_norway + norway_covid$weekly_death, lim = 13)$plot + ggtitle("Norway")
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig13a} Posterior predictive distribution of deaths in 2020 compared to mean deaths plus reported COVID-19 deaths"}
excess_plot_england_covid
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig13b} Posterior predictive distribution of deaths in 2020 compared to mean deaths plus reported COVID-19 deaths - continued"}
excess_plot_belgium_covid
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig13c} Posterior predictive distribution of deaths in 2020 compared to mean deaths plus reported COVID-19 deaths - continued"}
excess_plot_france_covid
```

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig13d} Posterior predictive distribution of deaths in 2020 compared to mean deaths plus reported COVID-19 deaths - continued"}
excess_plot_norway_covid
```

```{r}
# Total excess of deaths on top of covid reported deaths  
df_england <- plot_excessdeaths(posterior_chains_england_covid, mean_england + england_covid, lim = 14)$comp
df_england <- df_england %>%
  mutate(
    q0.05 = Simq0.05 - meanpluscovid,
    median = Simmedian - meanpluscovid,
    q0.95 = Simq0.95 - meanpluscovid
  ) %>%
  filter(week >= 11) %>%
  summarise(
    q0.05 = sum(q0.05),
    median = sum(median),
    q0.95 = sum(q0.95)
  )

df_belgium <- plot_excessdeaths(posterior_chains_belgium_covid, mean_belgium + belgium_covid$weekly_death, lim = 14)$comp
df_belgium <- df_belgium %>%
  mutate(
    q0.05 = Simq0.05 - meanpluscovid,
    median = Simmedian - meanpluscovid,
    q0.95 = Simq0.95 - meanpluscovid
  ) %>%
  filter(week >= 12) %>%
  summarise(
    q0.05 = sum(q0.05),
    median = sum(median),
    q0.95 = sum(q0.95)
  )

df_france <- plot_excessdeaths(posterior_chains_france_covid, mean_france + france_covid$weekly_death, lim = 14)$comp
df_france <- df_france %>%
  mutate(
    q0.05 = Simq0.05 - meanpluscovid,
    median = Simmedian - meanpluscovid,
    q0.95 = Simq0.95 - meanpluscovid
  ) %>%
  filter(week >= 11) %>%
  summarise(
    q0.05 = sum(q0.05),
    median = sum(median),
    q0.95 = sum(q0.95)
  )
```

Finally in Table \ref{tab:tab9}, we compute estimates of the total excess deaths on top of the expected number of deaths plus the reported COVID-19 deaths for Belgium, France and England and Wales, as well as a 95% credible interval. To do so, we sum the weekly estimates of excess deaths over the outbreak weeks for each country: 11 to 18 for France and England and Wales and 12 to 18 for Belgium. We find positive point estimates for all the countries. However, for Belgium the 95% credible interval includes null excess deaths and negative values. The point estimate of excess deaths represents 47% of reported COVID-19 deaths for England and Wales, 11% for France and 7% for Belgium.

```{r}
df_excess <- rbind(df_england, df_belgium, df_france)
df_excess <- data.frame(
  country = c("England and Wales", "Belgium", "France"),
  df_excess,
  reported = c(
    sum(england_covid),
    sum(belgium_covid$weekly_death),
    sum(france_covid$weekly_death)
  )
)
df_excess <- df_excess %>% mutate(diff = median / reported * 100)

kable(df_excess, "latex",
  booktabs = T,
  align = "c",
  digit = 0,
  col.names = c("Country", "Excess Q5%", "Excess Median", "Excess Q95%", "Reported COVID", "Difference %"),
  caption = "\\label{tab:tab9} Total excess deaths"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  )
```


```{r,eval=FALSE}
save.image("data_report.Rdata")
```


# Conclusion

Our results show that in France, Belgium and England and Wales there have been several weeks of higher than usual mortality in March and April 2020. This is not the case in Norway. We also find excess deaths on top of the expected number of deaths and the reported COVID-19 deaths in France and England and Wales. The reported COVID-19 deaths do not explain entirely the surge in deaths observed in those months. This is particularly true in England and Wales where the excess is particularly large, despite COVID-19 deaths figures for the country being more pessimistic than those of the other countries considered. On the contrary, Belgium has a small excess in deaths and reported COVID-19 explain well the observed surge in deaths.

These results suggest an under reporting of pandemics-related deaths in England & Wales and in France and a more accurate reporting in Belgium. However, further development of the methodology used here and of the scope of countries and weeks could help provide stronger evidence. In particular, we only considered data reported up to week 18 but the outbreak had not yet ended in England and Wales and later adjustments to COVID-19 related deaths were made in many countries. Adding more countries and integrating them all in a single hierarchical model could also help to obtain a fairer estimate of the relative risk. Another interesting development is that our model can be used to evaluate the impact of lockdown measures and country-wide risk factors by adding explanatory variables in the fixed and time structured components of relative risk.

# References

[1] https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv

[2] https://www.euronews.com/2020/04/06/a-little-flu-brazil-s-bolsonaro-playing-down-coronavirus-crisis

[3] https://www.ft.com/content/6bd88b7d-3386-4543-b2e9-0d5c6fac846c

[4] https://www.worldometers.info/coronavirus/#countries

[5] https://www.bbc.com/news/world-europe-52491210

[6] Statistics Norway, https://www.ssb.no/statbank/table/07995/

[7] Statbel, https://statbel.fgov.be/fr/nouvelles/mortalite-jusquau-3-mai

[8] Institut National de Statistiques et tudes conomiques, https://www.insee.fr/fr/statistiques/4487854, https://www.insee.fr/fr/information/4190491

[9] Office of National Statistics, https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/weeklyprovisionalfiguresondeathsregisteredinenglandandwales

# Annex

## Validation of excess deaths model

```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig9c} Validation of excess deaths model continued"}
val_plot_belgium
```


```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig9d} Validation of excess deaths model continued"}
val_plot_france
```


```{r}
df1 <- cbind(
  seq(1:18), plot_validation(posterior_chains_england, england_2020, lim = 14)$comp_med,
  plot_validation(posterior_chains_belgium, belgium_2020, lim = 13)$comp_med
)
df2 <- cbind(
  seq(1:18), plot_validation(posterior_chains_france, france_2020, lim = 13)$comp_med,
  plot_validation(posterior_chains_norway, norway_2020, lim = 13)$comp_med
)

kable(df1, "latex",
  booktabs = T,
  align = "c",
  digit = 0,
  col.names = c("week", rep(c("Obs. 2020", "Sim. Median"), 2)),
  caption = "\\label{tab:tab5} Values observed in 2020 and simulations median in excess deaths model"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  ) %>%
  add_header_above(c(" " = 1, "England and Wales" = 2, "Belgium" = 2))


kable(df2, "latex",
  booktabs = T,
  align = "c",
  digit = 0,
  col.names = c("week", rep(c("Obs. 2020", "Sim. Median"), 2)),
  caption = "\\label{tab:tab6} Values observed in 2020 and simulations median in excess deaths model"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  ) %>%
  add_header_above(c(" " = 1, "France" = 2, "Norway" = 2))
```

## Validation of excess deaths model on top of reported COVID-19 deaths


```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig10} Validation of excess deaths on top of the reported COVID-19 deaths model"}
val_plot_england_covid
```


```{r, echo = FALSE, eval = TRUE, fig.cap="\\label{fig:fig10b} Validation of excess deaths on top of the reported COVID-19 deaths model continued"}
val_plot_norway_covid
```


```{r}
df_covid1 <- cbind(
  seq(1:18), plot_validation(posterior_chains_england_covid, england_2020, lim = 14)$comp_med,
  plot_validation(posterior_chains_belgium_covid, belgium_2020, lim = 13)$comp_med
)
df_covid2 <- cbind(
  seq(1:18), plot_validation(posterior_chains_france_covid, france_2020, lim = 13)$comp_med,
  plot_validation(posterior_chains_norway_covid, norway_2020, lim = 13)$comp_med
)

kable(df_covid1, "latex",
  booktabs = T,
  align = "c",
  digit = 0,
  col.names = c("week", rep(c("Obs. 2020", "Sim. Median"), 2)),
  caption = "\\label{tab:tab7} Values observed in 2020 and simulations median in excess deaths on top of the reported COVID-19 deaths model"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  ) %>%
  add_header_above(c(" " = 1, "England and Wales" = 2, "Belgium" = 2))

kable(df_covid2, "latex",
  booktabs = T,
  align = "c",
  digit = 0,
  col.names = c("week", rep(c("Obs. 2020", "Sim. Median"), 2)),
  caption = "\\label{tab:tab8} Values observed in 2020 and simulations median in excess deaths on top of the reported COVID-19 deaths model"
) %>%
  kable_styling(
    latex_options = c("hold_position", "condensed")
  ) %>%
  add_header_above(c(" " = 1, "France" = 2, "Norway" = 2))
```
